---
params:
    root_directory: NULL
    input_seurat_object: NULL
    output_seurat_object: NULL
    cell_type_markers: NULL
    transcriptome_id_map: NULL
title: "NBIS Support #5568"
subtitle: "Integration and known cell type markers"
author: "Erik Fasterius"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        code_folding: hide
        toc: true
        toc_float:
            collapsed: true
---

```{r Setup}
# Knit options
knitr::opts_knit$set(root.dir = params$root_directory)

# Chunk options
knitr::opts_chunk$set(fig.width  = 12,
                      fig.height = 8,
                      fig.align  = "center",
                      message    = FALSE,
                      warning    = FALSE,
                      results    = "asis")

# Set seed
set.seed(42)
```

```{r Load packages}
suppressPackageStartupMessages({
    library("cowplot")
    library("Seurat")
    library("tidyverse")
})
```

```{r Data and variables}
# Read filtered data
s_obj <- readRDS(params$input_seurat_object)
# s_obj <- readRDS("results/seurat/01-quality_controls/seurat-qc.rds")

# Define number of PCs and variable features
npcs <- 50
nfeatures <- 3000
```

# Integration

We first want to integrate our two samples together using overlapping cell
types to help clustering and downstream analyses. This only affects cells that
are the same in the two samples; different cell types will still yield separate
clusters.

```{r Remove doublets, results = FALSE, include = FALSE, eval = FALSE}
# Estimate ideal PC neighbourhood size for computing pANN
# (proportion of artifical k nearest neighbours)
library("DoubletFinder")
sweep_res <- paramSweep_v3(s_obj, PCs = 1:npcs, sct = FALSE)
sweep_stats <- summarizeSweep(sweep_res, GT = FALSE)
sweep_results <- find.pK(sweep_stats)
max_bc <- max(sweep_results$BCmetric)
pK_estimate <- sweep_results[sweep_results$BCmetric == max_bc, "pK"]
pK_estimate <- as.numeric(as.character(pK_estimate))

# Define pANN threshold for singlet/doublet prediction
# (using estimates from 10X for loading 16000 cells yielding ~7.6% doublets)
n_exp <- round(0.076 * nrow(s_obj@meta.data))

# Adjust pANN threshold with estimated proportion of homotypic doublets
prop_homotypic <- modelHomotypic(s_obj@meta.data$seurat_clusters)
n_exp_adj <- round(n_exp * (1 - prop_homotypic))

# Run DoubletFinder with varying stringency criteria
s_obj <- doubletFinder_v3(s_obj,
                          PCs        = 1:npcs,
                          pN         = 0.25,
                          pK         = pK_estimate,
                          nExp       = n_exp,
                          reuse.pANN = FALSE,
                          sct        = FALSE)
```

```{r Integrate datasets}
# Split data into each original sample
s_obj_list <- SplitObject(s_obj, split.by = "orig.ident")

# Normalise and find highest variable features per dataset
for (i in 1:length(s_obj_list)) {
    s_obj_list[[i]] <- NormalizeData(s_obj_list[[i]], verbose = FALSE)
    s_obj_list[[i]] <- FindVariableFeatures(s_obj_list[[i]],
                                              selection.method = "vst",
                                              nfeatures        = nfeatures,
                                              verbose          = FALSE)
}

# Find integration anchors
anchors <- FindIntegrationAnchors(object.list = s_obj_list,
                                  dims        = 1:npcs,
                                  reduction   = "cca")

# Integrate the samples with the identified anchors
s_obj_integrated <- IntegrateData(anchorset      = anchors,
                                  dims           = 1:npcs,
                                  new.assay.name = "CCA")

# Run scaling, regression, PCA and find nearest neighbours
to_regress <- c("percent_mito", "percent_ribo", "percent_hsp", "nFeature_RNA",
                "S.Score", "G2M.Score")
s_obj_integrated <- ScaleData(s_obj_integrated, vars.to.regress = to_regress)
s_obj_integrated <- RunPCA(s_obj_integrated, npcs = npcs, verbose = FALSE)
s_obj_integrated <- FindNeighbors(s_obj_integrated, dims = 1:npcs)
```

```{r Find optimal cluster resolution, eval = FALSE, include = FALSE}
# Find optimal cluster resolution
for (resolution in seq(0.1, 2.0, by = 0.1)) {
    s_obj_temp <- FindClusters(s_obj_integrated,
                               resolution = resolution,
                               verbose    = FALSE)
    s_obj_temp <- RunUMAP(s_obj_temp, dims = 1:npcs)
    print(DimPlot(s_obj_temp,
            reduction = "umap",
            group.by  = "seurat_clusters",
            label     = TRUE) +
        ggtitle(paste("Resolution =", resolution)) +
        NoAxes())
}
```

```{r Find clusters at optimal resolution and run UMAP}
s_obj_integrated <- FindClusters(s_obj_integrated,
                                 resolution = 0.7,
                                 verbose    = FALSE)
s_obj_integrated <- RunUMAP(s_obj_integrated, dims = 1:npcs)
```

```{r Plot integrated clusters}
# Plot clusters
plot_grid(ncol = 2,
    DimPlot(s_obj, reduction = "umap", group.by = "orig.ident") +
        NoAxes() +
        ggtitle("UMAP raw data"),
    DimPlot(s_obj_integrated, reduction = "umap", group.by = "orig.ident") +
        NoAxes() +
        ggtitle("UMAP integrated data")
)
```

```{r Plot integrated features, fig.height = 16}
features <- c("nCount_RNA", "nFeature_RNA", "percent_mito",
              "percent_ribo", "S.Score", "G2M.Score")
FeaturePlot(s_obj_integrated, reduction = "umap", features = features)
```

# Marker expression

Here we simply map and plot the expression of the previously known cell type
markers in each cell and cluster.

```{r Get cell type markers}
# Read cell type markers
cell_type_markers <- read.table(params$cell_type_markers,
                                sep              = ",",
                                header           = TRUE,
                                stringsAsFactors = FALSE)

# Read transcriptome ID mappings
marker_mapping <- read.table(params$transcriptome_id_map,
                             sep              = "\t",
                             header           = TRUE,
                             stringsAsFactors = FALSE)
marker_mapping$original_id <- gsub("_", "-", marker_mapping$original_id)

# Merge to get original IDs used in the Seurat object
markers <- merge(cell_type_markers,
                 marker_mapping,
                 by.x = "NCBI_accession",
                 by.y = "ncbi_id")
```

```{r Plot cell type markers}
# Re-order clusters
num_clusters <- length(levels(s_obj_integrated)) - 1
s_obj_integrated$seurat_clusters <-
    factor(s_obj_integrated$seurat_clusters,
           levels = c(0:num_clusters))
levels(s_obj_integrated) <- c(0:num_clusters)

# Plot clusters
DimPlot(s_obj_integrated,
        reduction = "umap",
        label     = TRUE,
        group.by  = "seurat_clusters") +
    NoAxes()

# Plot marker expression
FeaturePlot(s_obj_integrated,
            reduction = "umap",
            order     = TRUE,
            features  = markers$original_id)
VlnPlot(s_obj_integrated,
        assay    = "RNA",
        features = markers$original_id)
```

The above figures are violin-plots of cell type marker expression, but since
almost all cells in all clusters show a zero or near-zero expression of the
markers it looks like flat distributions around zero, with some outlier cells
expressing the markers scattered at low (but non-zero) expression. We can,
however, see a slight pattern in the expression of the `comp36642-c0-seq6`
transcript, in that it is slightly more expressed in cluster 4, 5, 6 and 10 -
most of the cells in these clusters do not, however, express this transcript at
all, but it is the only discernible pattern here.

Below we also plot some additional markers that arose as interesting during the
analyses; most features are not highly expressed in any clusters, but a small
number of them stand out as having one or a few clusters with higher expression
than others: PACIHC, PXN, TGA1 and VMO1.

```{r Plot additional markers, fig.height = 48}
# Add manual features from group
additional_markers <- c("A2m", "CHF", "CP", "Dclk1", "Dclk2", "Drgx",
                        "Dscam2", "Dscaml1", "KPI-46742", "KPIc", "LGBP",
                        "MAP2", "MAS", "MBL", "Nmdar1", "PACIHC", "PAPI",
                        "PLCSP", "PPA", "ProPO", "PVF", "PVR", "PXN",
                        "SGC-KPI", "SOD", "SPH1", "SPH2", "TGA1", "TGA2",
                        "VMO1")

# Plot marker expression
FeaturePlot(s_obj_integrated,
            reduction = "umap",
            order     = TRUE,
            ncol      = 2,
            features  = additional_markers)
VlnPlot(s_obj_integrated,
        assay    = "RNA",
        ncol     = 2,
        features = additional_markers)
```

```{r Save RDS object}
saveRDS(s_obj_integrated, file = params$output_seurat_object)
```

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
