---
params:
    root_directory: NULL
    input_seurat_object: NULL
title: "NBIS Support #5568"
subtitle: "Features and cell types"
author: "Erik Fasterius"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        code_folding: hide
        toc: true
        toc_float:
            collapsed: true
---

```{r Setup}
# Knit options
knitr::opts_knit$set(root.dir = params$root_directory)

# Chunk options
knitr::opts_chunk$set(fig.width  = 14,
                      fig.height = 7,
                      fig.align  = "center",
                      message    = FALSE,
                      warning    = FALSE,
                      results    = "asis")

# Set seed
set.seed(42)

# Load packages
suppressPackageStartupMessages({
    library("cowplot")
    library("ggplot2")
    library("scales")
    library("Seurat")
    library("slingshot")
    library("tidyverse")
    library("tradeSeq")
})

# Variables
nfeatures <- 3000
npcs <- 50

# Parallelisation
BiocParallel::register(BiocParallel::SerialParam())
```

```{r Read data}
# Seurat object
s_obj <- readRDS(params$input_seurat_object)
# s_obj <- readRDS("results/seurat/04-features-and-celltypes/seurat-features-and-celltypes.rds")
DefaultAssay(s_obj) <- "RNA"
```

Trajectory inference can only be run on cells where you expect to find some kind
of dimension of change, *e.g.*  differentiation. We thus start with subsetting
the data to each of the "clouds" and perform trajectory analysis for each cloud.

```{r Integration functions}
# Function to re-integrate and re-cluster data
integrate_data <- function(s_obj,
                           split_by   = "orig.ident",
                           resolution = 0.8,
                           dims       = 50,
                           nfeatures  = 3000) {

    # Split data into each original sample
    s_obj_list <- SplitObject(s_obj, split.by = split_by)

    # Normalise and find highest variable features per dataset
    for (i in 1:length(s_obj_list)) {
        s_obj_list[[i]] <- NormalizeData(s_obj_list[[i]], verbose = FALSE)
        s_obj_list[[i]] <- FindVariableFeatures(s_obj_list[[i]],
                                                selection.method = "vst",
                                                nfeatures        = nfeatures,
                                                verbose          = FALSE)
    }

    # Find integration anchors
    anchors <- FindIntegrationAnchors(object.list = s_obj_list,
                                      dims        = 1:dims,
                                      reduction   = "cca")

    # Integrate the samples with the identified anchors
    s_obj_integrated <- IntegrateData(anchorset      = anchors,
                                      dims           = 1:npcs,
                                      new.assay.name = "CCA")

    # Run scaling, regression, PCA and find nearest neighbours
    to_regress <- c("percent_mito", "percent_ribo", "nFeature_RNA", "nCount_RNA")
    s_obj_integrated <- ScaleData(s_obj_integrated, vars.to.regress = to_regress)
    s_obj_integrated <- RunPCA(s_obj_integrated, npcs = dims, verbose = FALSE)
    s_obj_integrated <- FindNeighbors(s_obj_integrated, dims = 1:dims)

    # Cluster integrated data
    s_obj_integrated <- FindClusters(s_obj_integrated,
                                     resolution = resolution,
                                     verbose    = FALSE)
    s_obj_integrated <- RunUMAP(s_obj_integrated,
                                dims = 1:dims)

    # Set default assay to RNA
    DefaultAssay(s_obj_integrated) <- "RNA"

    # Return integrated object
    return(s_obj_integrated)
}
```

```{r Trajectory inference functions}
# Function to get colours for trajectory plots
get_colours_for_trajectory <- function(grouping_values) {

    # Get palette for the specified grouping values
    categories <- sort(unique(grouping_values))
    palette <- setNames(hue_pal()(length(categories)), categories)

    # Return grouping values with corresponding colours
    return(palette[grouping_values])
}

# Function for getting Slingshot lineages
infer_trajectories <- function(s_obj,
                               grouping_variable = "seurat_clusters",
                               start_cluster     = NULL,
                               reduction         = "umap",
                               approx_points     = 300) {

    # Get lineages
    dim_red <- s_obj@reductions[[reduction]]@cell.embeddings
    grouping_values <- factor(s_obj@meta.data[, grouping_variable])
    lineages <- getLineages(data          = dim_red,
                            start.clus    = start_cluster,
                            clusterLabels = grouping_values)

    # Get curves
    curves <- getCurves(lineages,
                        approx_points = approx_points,
                        allow.breaks  = FALSE,
                        thresh        = 0.01,
                        stretch       = 0.8,
                        shrink        = 0.99)

    # Get colour palette for the specified grouping values
    categories <- sort(unique(grouping_values))
    palette <- setNames(hue_pal()(length(categories)), categories)

    # Initialise three-column plot
    par(mfrow = c(1, 3))

    # Plot clusters
    plot(dim_red[, 1:2],
         col = palette[grouping_values],
         cex = .5,
         pch = 16)
    for(ii in levels(grouping_values)) {
      text(mean(dim_red[grouping_values == ii, 1]),
           mean(dim_red[grouping_values == ii, 2]),
           labels = ii,
           font   = 2)
    }

    # Plot lineages
    plot(dim_red[, 1:2],
         col = palette[grouping_values],
         cex = .5,
         pch = 16)
    lines(SlingshotDataSet(lineages),
          lwd = 3,
          col = "black")

    # Plot curves
    plot(dim_red[, 1:2],
         col = palette[grouping_values],
         cex = .5,
         pch = 16)
    lines(SlingshotDataSet(curves),
          lwd = 3,
          col = "black")

    # Return the curves
    return(curves)
}
```

```{r Trajectory DEA functions}
# Function to fit generalised additive model
fit_gam <- function(s_obj,
                    curves,
                    plot_knots        = FALSE,
                    nfeatures         = 2000,
                    hvg_assay         = "CCA",
                    grouping_variable = "seurat_clusters") {

    # Find highest variable features
    s_obj <- FindVariableFeatures(s_obj,
                                  assay     = hvg_assay,
                                  nfeatures = nfeatures)
    hvg <- VariableFeatures(s_obj, assay = hvg_assay)

    # Fit GAM
    counts <- as.matrix(s_obj@assays$RNA@counts)
    sce <- fitGAM(counts = counts,
                  genes  = hvg,
                  sds    = curves)

    # Plot knots (if applicable)
    if (plot_knots) {
        grouping_values <- factor(s_obj@meta.data[, grouping_variable])
        gg <- plotGeneCount(curves,
                            models   = sce,
                            counts   = counts,
                            clusters = grouping_values)
        show(gg)
    }

    # Return fitted model
    return(sce)
}

# Function to plot trajectory DEGs
plot_trajectory_deg <- function(feature_id,
                                sce,
                                s_obj,
                                curves,
                                grouping_variable = "seurat_clusters") {
    # Plot knots
    grouping_values <- factor(s_obj@meta.data[, grouping_variable])
    counts <- as.matrix(s_obj@assays$RNA@counts)
    gg1 <- plotGeneCount(curves,
                         counts,
                         clusters = grouping_values,
                         models   = sce,
                         gene     = feature_id) +
        theme(legend.position = "none") +
        labs(x = "UMAP_1", y = "UMAP_2") +
        ggtitle(feature_id)

    # Plot gene smoothers
    gg2 <- plotSmoothers(sce, counts, gene = feature_id) +
        labs(colour = "Lineage")
    gg <- plot_grid(gg1, gg2)

    # Return graphical object
    return(gg)
}

# Function to get trajectory DEGs
get_trajectory_degs <- function(sce,
                                s_obj,
                                curves,
                                output_file,
                                FUN          = associationTest,
                                l2fc         = log2(2),
                                p_val_cutoff = 0.05,
                                top_n        = 10) {

    # Get trajectory DEGs
    res <- FUN(sce, l2fc = l2fc)

    # Filter DEGs
    degs <- res %>%
        mutate(feature = row.names(.)) %>%
        select(feature, everything()) %>%
        filter(pvalue < p_val_cutoff) %>%
        arrange(desc(waldStat))

    # Plot top n DEGs
    gg_list <- list()
    for (feature_id in row.names(degs)[1:top_n]) {
        gg <- plot_trajectory_deg(feature_id, sce, s_obj, curves)
        gg_list[[length(gg_list) + 1]] <- gg
    }
    gg <- plot_grid(plotlist = gg_list, ncol = 1)
    show(gg)

    # Write to file
    write.table(degs,
                file      = output_file,
                row.names = FALSE,
                sep       = "\t")

    # Return list of DEGs
    return(degs)
}
```

# Left cloud

```{r Subset to LC cells}
# Subset cells and re-integrate
s_obj_LC <- subset(s_obj, subset = cell_type %in% c("g1",
                                                    "g2",
                                                    "g3",
                                                    "sg1",
                                                    "sg2"))
```

```{r Find ideal LC trajectories, fig.width = 21, eval = FALSE, include = FALSE}
# Run trajectory analyses for each resolution
resolutions <- seq(0.1, 1.0, 0.1)
for (resolution in resolutions) {
    s_obj_LC_current <- integrate_data(s_obj_LC, resolution = resolution)
    trajectories <- infer_trajectories(s_obj_LC_current)
}
```

```{r Integrate LC data}
# Integrate data
s_obj_LC <- integrate_data(s_obj_LC, resolution = 0.2)

# Plot cell types
gg1 <- DimPlot(s_obj_LC, group.by = "seurat_clusters", label = TRUE)
gg2 <- DimPlot(s_obj_LC, group.by = "cell_type", label = TRUE)
gg <- plot_grid(gg1, gg2, ncol = 2)
gg
```

```{r Infer LC trajectories, fig.width = 21}
curves_LC <- infer_trajectories(s_obj_LC, "seurat_clusters")
```

```{r Fit LC GAM}
sce_LC <- fit_gam(s_obj_LC, curves_LC)
```

## DEGs across trajectories

Differential expression across trajectories can be calculated in several ways.
As the LC trajectories don't have any branches, we'll simply calculated DEGs
across the trajectory in general, *i.e.* changes across pseudotime. We'll plot
the top 10 significantly differentially expressed genes.

```{r Get LC trajectory DEGs, fig.height = 70, fig.width = 14}
# DEGs associated with change across trajectories
degs_LC <- get_trajectory_degs(sce         = sce_LC,
                               s_obj       = s_obj_LC,
                               curves      = curves_LC,
                               output_file = "degs-LC-trajectory-overall.tsv",
                               FUN         = associationTest)
```

```{r LC cleanup}
rm(sce_LC, curves_LC, s_obj_LC)
invisible(gc())
```

# Right cloud

```{r Subset to RC cells}
# Subset cells and re-integrate
s_obj_RC <- subset(s_obj, subset = cell_type %in% c("pro_hemocyte_1",
                                                    "pro_hemocyte_2",
                                                    "progenitor",
                                                    "immune_cell",
                                                    "unknown",
                                                    "PASIHC"))
```

```{r Find ideal RC trajectories, fig.width = 21, eval = FALSE, include = FALSE}
# Run trajectory analyses for each resolution
resolutions <- seq(0.1, 1.0, 0.1)
for (resolution in resolutions) {
    s_obj_RC_current <- integrate_data(s_obj_RC, resolution = resolution)
    trajectories <- infer_trajectories(s_obj_RC_current)
}
```

```{r Integrate RC data}
# Integrate data
s_obj_RC <- integrate_data(s_obj_RC, resolution = 0.2)

# Plot cell types
gg1 <- DimPlot(s_obj_RC, group.by = "seurat_clusters", label = TRUE)
gg2 <- DimPlot(s_obj_RC, group.by = "cell_type", label = TRUE)
gg <- plot_grid(gg1, gg2, ncol = 2)
gg
```

```{r Infer RC trajectories, fig.width = 21}
curves_RC <- infer_trajectories(s_obj_RC, "seurat_clusters",
                                start_cluster = "3")
```

```{r Fit RC GAM}
sce_RC <- fit_gam(s_obj_RC, curves_RC)
```

## DEGs across trajectories

We'll start with DEGs associated with change across trajectories in general for
the RC trajectories, the same as for the LC trajectories.

```{r Get RC trajectory DEGs, fig.height = 70, fig.width = 14}
# DEGs associated with change across trajectories
degs_RC <- get_trajectory_degs(sce         = sce_RC,
                               s_obj       = s_obj_RC,
                               curves      = curves_RC,
                               output_file = "degs-RC-trajectory-overall.tsv",
                               FUN         = associationTest)
```

## DEGs between end points

Since the RC trajectories are branched, we can additionally calculated DEGs for
different ways across the branches. We'll start with DEGs that differ between
the end points of the two branches.

```{r Get RC end-point DEGs, fig.height = 70, fig.width = 14}
# DEGs associated with change across trajectories
degs_RC <- get_trajectory_degs(sce         = sce_RC,
                               s_obj       = s_obj_RC,
                               curves      = curves_RC,
                               output_file = "degs-RC-trajectory-endpoints.tsv",
                               FUN         = diffEndTest)
```

## DEGs between branching points

Let's also find DEGs that are differing at the branching points, rather at the
end points.

```{r Get RC branching-point DEGs, fig.height = 70, fig.width = 14}
# DEGs associated with change across trajectories
degs_RC <- get_trajectory_degs(sce         = sce_RC,
                               s_obj       = s_obj_RC,
                               curves      = curves_RC,
                               output_file = "degs-RC-trajectory-branches.tsv",
                               FUN         = earlyDETest)
```

<!-- Remove whitespace at end of document due to using floating ToC -->
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
